---
title: "Debugging models with posterior predictive checks"
author: ""
date: ""
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
```

```{r load-packages, message=FALSE, warning=FALSE}
library("ggplot2")
library("gridExtra")
library("bayesplot")
library("rstan")
```

```{r options}
options(mc.cores = parallel::detectCores())
theme_set(bayesplot::theme_default(y_lab = FALSE))
```


### Load and look at the data

```{r poisson-data}
source("count-data.R")
print(N)
print(y)
```

Make a plot of `y`

```{r plot-y}


```

Compare to random draws from a Poisson with the same mean

```{r compare}


```


### Fit basic Poisson model
```{r fit, results="hide", message=FALSE, warning=FALSE}


```

### Graphical posterior predictive checks

Extract `y_rep` draws from the fitted model object

```{r y_rep}


# what's the structure of y_rep object?
```

Compare histogram of `y` to histograms of several `y_rep`s

```{r ppc-hist}


```

Compare proportion of zeros in `y` and the distribution of that proportion 
over the `y_rep` replications

```{r prop-zero}
prop_zero <- # function to compute proportion of zeros

  
```

Check the proportion greater than `max(y)`

```{r prop-gt-max-y}
print(max(y))
prop_gt_max_y <- # function to compute proportion > max(y)

  
```

### Fit Poisson "hurdle" model with truncation

```{r fit2, results="hide", message=FALSE, warning=FALSE}
fit2 <- 
```

### Posterior predictive checks again

```{r ppc-2}

```
